### 2016-05-13 text/Chapter08.tex
Edit thesis conclusions (Chapter 9)

\chapter{Related and future work}
\label{ch:further}

@@

>>>>

\chapter{Further work}
\label{ch:further}

@@

% =================================================================================================
%
% 	 #####                 #     #                                                #
% 	 #                     #     #                                                #
% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
%
% =================================================================================================

\section{Further work}
\label{sec:conc-further}

Most of the further work has been discussed throughout the thesis, so this section
serves mainly as an overview. We follow the same structure as when discussing pathways to
coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
to coeffect systems, categorical semantics of coeffects, and connections with substructural type
systems and bunched logics.

\paragraph{Syntactic coeffect systems.}
The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
properties of coeffect systems (and explain how coeffect system differ from better-known effect
systems), but it is hardly sufficient for a realistic programming language.

Further work is to extend the coeffect systems presented here to a full programming language. A
useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
similar development for effect systems, adding conditionals, recursion and polymorphism.

Coeffect annotations in context-aware programming languages can contain significant amount
of information. Thus, the programming language also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For
\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where
\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.

\paragraph{Language semantics.}

As discussed in Section~\ref{sec:path-sem}, comonads can be used to define the semantics of a
programming language in two ways. The first, ``language semantics'', approach is to use a
single comonad to describe the semantics of a specific context-aware property. The second,
``meta-language'' approach is to extend the language with explicit constructs representing the
operations of the underlying comonad.

The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
guide for the development of the type systems. Further work includes more precise treatment of
the categorical structure of indexed comonads. This has partly been done for flat coeffects by
Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
work is to develop a similar categorical model for the unified coeffect calculus, possibly using
the categorical notion of containers \cite{types-containers}.

We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
Developing this technique further could unify coeffect systems with the work on Contextual Modal
Type-Theory (CMTT) \cite{logic-cmtt} and allow other interesting applications of coeffects such
as meta-programming \cite{logic-cmtt} and distributed computing with explicit modalities \cite{app-distributed-ml5}.

\paragraph{Substructural and bunched logics.}

In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
context} and to \emph{individual variables}, respectively. We later unified the two systems
in Section~\ref{sec:unified-unified}.

An intriguing question is whether coeffects can generalize \emph{bunched typing} \cite{substruct-bunched},
which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather
than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as
bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.

\paragraph{Applications.}
The last direction for further development is developing practical programming languages or
libraries based on the theory presented in this thesis. We believe that the most useful approach
is to extend programming languages with a reusable coeffect system and allow library developers
define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
``do'' notation.

Another option is to use coeffects just as the underlying theory of a single-purpose programming
language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
we covered a number of concrete problems that developers are facing today. Each of those would
presents an important further work. The most important practical challenges are cross-compilation
(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
and security properties.

### 2016-05-13 text/Chapter09.tex
Edit thesis conclusions (Chapter 9)

\chapter{Conclusions and further work}
\label{ch:conclusions}

@@

In this chapter, we give a brief summary of the technical development completed in this thesis
(Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
common patterns and captures those using three \emph{coeffect calculi}. Next, we give an overview
of further work (Section~\ref{sec:conc-further}), often referring to longer discussion in earlier
chapters. Finally, Section~\ref{sec:conc-conclusions} concludes the thesis.

% ==================================================================================================
%
%       ###
%      #   #
%      #      #   #  ## #   ## #    ###   # ##   #   #
%       ###   #   #  # # #  # # #      #  ##  #  #   #
%          #  #   #  # # #  # # #   ####  #      #  ##
%      #   #  #  ##  # # #  # # #  #   #  #       ## #
%       ###    ## #  #   #  #   #   ####  #          #
%                                                #   #
%                                                 ###
% =================================================================================================

\section{Summary}
\label{sec:conc-summary}

Modern computer programs run in rich and diverse environments or \emph{contexts}. The richness means
that environments provides additional resources and capabilities that may be accessed by the program.
The diversity means that programs often need to run in multiple different environments, such as mobile
phones, servers, web browsers or even on the GPU. In this thesis, we present the foundations for
building programming languages that simplify writing software for such rich and diverse environments.

@@

@@

Our focus is on the syntactic properties of the calculus -- we give a type system
(Section~\ref{sec:flat-calculus}) and discuss equational theory of the calculus (Section~\ref{sec:flat-syntax}).
In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
the type of an expression, but we identify two conditions when this is the case -- this gives us
a basis for operational semantics of our calculi for liveness and implicit parameters.

The design of the type system is further validated by categorical semantics
(Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of
\emph{flat indexed comonad} -- a structure based on a categorical dual of monads.

@@

\paragraph{Unified calculus.}

The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
thesis. In Chapter~\ref{ch:further}, we discuss the similarities and differences between the two.
Although the calculi are worth considering separately (as they have different equational properties),
they can be presented in a uniform way. In Section~\ref{sec:further-unified}, we present the
\emph{unified coeffect calculus}, which can be instantiate to track both per-variable and
whole-context properties. The key trick is to parameterize the calculus by the \emph{shape} of
a context, which can be a vector (corresponding to the vector of free variables) or a singleton
container (attaching one annotation to the whole context).

% =================================================================================================
%
% 	 #####                 #     #                                                #
% 	 #                     #     #                                                #
% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
%
% =================================================================================================

\section{Further work}
\label{sec:conc-further}

Most of the further work has been discussed throughout the thesis, so this section
serves mainly as an overview. We follow the same structure as when discussing pathways to
coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
to coeffect systems, categorical semantics of coeffects, and connections with substructural type
systems and bunched logics.

% -------------------------------------------------------------------------------------------------

\paragraph{Syntactic coeffect systems.}
The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
properties of coeffect systems (and explain how coeffect system differ from better-known effect
systems), but it is hardly sufficient for a realistic programming language.

Further work is to extend the coeffect systems presented here to a full programming language. A
useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
similar development for effect systems, adding conditionals, recursion and polymorphism.

Coeffect annotations in context-aware programming languages can contain significant amount
of information. Thus, the programming language also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For
\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where
\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.

% -------------------------------------------------------------------------------------------------

\paragraph{Language semantics.}

As discussed in Section~\ref{sec:path-sem}, comonads can be used to define the semantics of a
programming language in two ways. The first, ``language semantics'', approach is to use a
single comonad to describe the semantics of a specific context-aware property. The second,
``meta-language'' approach is to extend the language with explicit constructs representing the
operations of the underlying comonad.

The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
guide for the development of the type systems. Further work includes more precise treatment of
the categorical structure of indexed comonads. This has partly been done for flat coeffects by
Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
work is to develop a similar categorical model for the unified coeffect calculus, possibly using
the categorical notion of containers \cite{types-containers}.

We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
Developing this technique further could unify coeffect systems with the work on Contextual Modal
Type-Theory (CMTT) \cite{logic-cmtt} and allow other interesting applications of coeffects such
as meta-programming \cite{logic-cmtt} and distributed computing with explicit modalities \cite{app-distributed-ml5}.

% -------------------------------------------------------------------------------------------------

\paragraph{Substructural and bunched logics.}

In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
context} and to \emph{individual variables}, respectively. We later unified the two systems
in Section~\ref{sec:unified-unified}.

An intriguing question is whether coeffects can generalize \emph{bunched typing} \cite{substruct-bunched},
which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather
than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as
bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.

% -------------------------------------------------------------------------------------------------

\paragraph{Applications.}
The last direction for further development is developing practical programming languages or
libraries based on the theory presented in this thesis. We believe that the most useful approach
is to extend programming languages with a reusable coeffect system and allow library developers
define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
``do'' notation.

Another option is to use coeffects just as the underlying theory of a single-purpose programming
language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
we covered a number of concrete problems that developers are facing today. Each of those would
presents an important further work. The most important practical challenges are cross-compilation
(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
and security properties.

% =================================================================================================
%
% 	  ###                         ##                    #
% 	 #   #                         #
% 	 #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###
% 	 #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #
% 	 #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###
% 	 #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      #
% 	  ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####
%
% =================================================================================================

>>>>

\chapter{Conclusions}
\label{ch:conclusions}

@@

In this chapter, we give a brief overview of the technical contributions of this thesis
(Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
common patterns and captures those using two \emph{coeffect calculi}. It gives formal semantics
using categorical foundations and uses it as a basis for prototype implementation. Finally,
Section~\ref{sec:conc-conclusions} concludes the thesis.

% ==================================================================================================
%
%    #######
%    #     # #    # ###### #####  #    # # ###### #    #
%    #     # #    # #      #    # #    # # #      #    #
%    #     # #    # #####  #    # #    # # #####  #    #
%    #     # #    # #      #####  #    # # #      # ## #
%    #     #  #  #  #      #   #   #  #  # #      ##  ##
%    #######   ##   ###### #    #   ##   # ###### #    #
%
% =================================================================================================

\section{Overview of contributions}
\label{sec:conc-summary}

As observed in Chapter~\ref{ch:intro}, modern computer programs run in rich and diverse environments
or \emph{contexts}. The richness means that environments provides additional resources and
capabilities that may be accessed by the program. The diversity means that programs often need to
run in multiple different environments, such as mobile phones, servers, web browsers or even on the
GPU. In this thesis, we present the foundations for building programming languages that simplify
writing software for such rich and diverse environments.

@@

% --------------------------------------------------------------------------------------------------
~

@@

We define a type system for the calculus (Section~\ref{sec:flat-calculus}),
discuss how to resolve the ambiguity inherent in context-aware programming for sample
languages we consider (Section~\ref{sec:flat-unique}) and study equational properties of the
calculus (Section~\ref{sec:flat-syntax}). In the flat coeffect calculus, $\beta$-reduction and
$\eta$-expansion do not generally preserve the type of an expression, but we identify two
conditions when this is the case.

@@

\paragraph{Semantics and safety.}

Both of the coeffect calculi are parameterized and can be instantiated to obtain a language with
a type system that tracks concrete notion of context dependence. The coeffect calculus can be seen
as a framework for building concrete domain-specific context-aware programming languages. In
addition to the type system, the framework also provides a way for defining the semantics of
concrete domain-specific context-aware languages, guides their implementation and simplifies
safety proofs.

This is done using a \emph{comonadically-inspired} semantics. We generalize the category-theoretical
dual of monad to \emph{indexed comonad} and use it to define the semantics of context-aware
programs for both the flat coeffect calculus (Chapter~\ref{ch:semantics}) and the structural coeffect
calculus (Section~\ref{sec:struct-semantics}). Due to the ambiguity inherent in contextual properties,
we define the semantics over a typing derivation, but we also give a domain-specific algorithm for
choosing a unique typing derivation for each of our sample languages (Section~\ref{sec:flat-unique}
and Section~\ref{sec:struct-unique}).

We use comonads in a syntactic way, following the example of Haskell's use of monads and treat
it as a \emph{translation} from source context-aware programming language to a simple target
functional language. The target language includes uninterpreted comonadically-inspired primitives
whose concrete reduction rules need to be provided for each concrete context-aware deomain-specific
language. We give concrete definitions and prove type safety for three sample context-aware
languages (Section~\ref{sec:semantics-proofs} and Section~\ref{sec:struct-transl}). Together with
well-typedness of our translation, these guarantee that ``well-typed context-aware programs do
not go wrong''.

\paragraph{Interactive essay.}

A part of the thesis is an interactive web-based essay (\url{http://tomasp.net/coeffects}),
which implements three simple context-aware languages. In addition to more traditional prototype
language implementations, the essay provides a number of novel features. As usual, it lets users
write and evaluate sample programs, but it also lets them explore the details of the coeffect
theory -- this includes an interactive way of exploring typing derivations and program semantics.
The essay also provides a number of case studies that highlight interesting aspects of the theory.

The implementation (Chapter~\ref{ch:impl}) serves two purposes. First, it links together all parts
of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the two coeffect calculi
to build a language framework that is then instantiated to implement three concrete context-aware
languages. The  framework provides the type system and translation (based on the
comonadically-inspired semantics) that is used for program execution. This shows the practical
benefits of using the theory of coeffects as a basis of real-world programming languages.

Second, the implementation uses novel kind of medium (Section~\ref{sec:impl-essay}) to make our work
\emph{accessible} and \emph{explorable}. Building a production-ready programming language is outside
the scope of the  thesis, so instead, our goal is to explain the theory and inspire authors of
real-world programming languages to include support for context-aware programming, ideally using
coeffects as a sound foundation. For this reason, we make the implementation accessible over web
without requiring installation of specialized software and we make it explorable to encourage
active reading.

% =================================================================================================
%
%     #####
%    #     # #    # #    # #    #   ##   #####  #   #
%    #       #    # ##  ## ##  ##  #  #  #    #  # #
%     #####  #    # # ## # # ## # #    # #    #   #
%          # #    # #    # #    # ###### #####    #
%    #     # #    # #    # #    # #    # #   #    #
%     #####   ####  #    # #    # #    # #    #   #
%
% =================================================================================================
